Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-03-29 RNARedPrint-Bioinfo2018.tex
Restated optimization

More complex \Def{Loop-based}
 energy models ---e.g.~the Turner model,
which among others adds energy terms for special loops and dangling
ends---are encoded as straightforward extensions. An interesting
stripped-down variant of the nearest neighbor model is the
\Def{stacking energy model}. This model assigns non-zero energy
contributions only to stacks, i.e. interior loops with closing base
pair $(i,j)$ and inner base pair $(i+1,j-1)$.

>>>>

More complex \Def{loop-based}
 energy models ---e.g.~the Turner model,
which among others includes energy terms for special loops and dangling
ends---can also be encoded as straightforward extensions. An interesting
stripped-down variant of the nearest neighbor model is the
\Def{stacking energy model}. This model assigns non-zero energy
contributions only to stacks, i.e. interior loops with closing base
pair $(i,j)$ and inner base pair $(i+1,j-1)$.

### 2018-03-29 supplement.tex
Restated optimization

\textcolor{red}{\textbf{SW: Simpler alternative to
    Sec.~\ref{sec:first-improvedComplexity}. Note also that the use of
    variables in Sec.~\ref{sec:first-improvedComplexity} conflicts
    with the (recent) specialization of constraint networks to RNA design
    networks.}}

@@

The correctness of the modified algorithm still follows from an
induction argument: in this way, the message computation over one node
evaluates messages from its children only at cc-valid partial
sequences. The result of this computation is a message, which corresponds 
to the partition function, restricted to cc-valid partial sequences. 
Since invalid sequences have infinite energy, they do not contribute to 
the partition function, and the restricted partition function coincides 
with the initial one.

Remarkably, this restriction drastically improves the time
complexity. While the original algorithm enumerates all partial
sequences $\chi(v):=\substitute{\val}{\val'}$ such that
$\val\in\partseqs(\separator{u}{v})$ and
$\val'\in\partseqs(\difference{u}{v})$, the modified algorithm
only considers cc-valid sequences for $\chi(v)$ the nodes of $v$.
Namely,

For each connected component having at least a position $i$ in $\chi(v)$, 
the algorithm considers assigning any possible nucleotide $x\in \Sigma$ to $i$ and restricts
 the assignment of  for 
any remaining position $j$ of the same connected component, it restricts the domain of $j$ to base-pairs that are consistent with the parity .

Let $c(v)$ be the number of connected components involving positions
in $\chi(v)$; then, the first stage enumerates $2^{c(v)}$
combinations. For each such combination, the second stage enumerates
only $2^{|\chi(v)|}$ partial sequences, since there remain only two
possible nucleotide choices for each position. All message values for
node $v$ are thus computed in time
$\Theta(2^{c(v)+|\chi(v)|})$. Consequently, the total time complexity
of the modified algorithm is $$O(nk2^{\hat w}),$$ where $n$ is the RNA
length, $k$ is the number of structures, and
$\hat w := \max_{v\in T} c(v)+|\chi(v)|$.
%
From this directly follows the slightly looser complexity
bound $$O(nk2^{w+1}2^{c})$$ in terms of treewidth $w$ and the maximum
number $c$ of connected components represented in a node of the tree
decomposition.  Notably, $\hat w$, as well as $w+1+c$, can not exceed
$2(w+1)$, since $c\leq w+1$. Consequently the modified algorithm does
never increase the original complexity.

Finally, we remark that even stronger time savings could be possible
in practice, since cc-valid partial sequences can still violate
complementarity constraints, e.g. by assigning C and A to positions in
different sets of a partition, thus satisfying the bi-partition
constraints, where the positions base pair directly, rendering the
partial sequence invalid. Moreover, applications of the sampling
framework, can introduce additional constraints that further reduce
the number of valid partial subsequences. However, exploiting all such
constraints, in a complete and general way, likely causes significant
overhead in an implementation and does not straightforwardly improve
the theoretical bound.

@@

We trained parameters for two RNA energy models to approximate the
Turner energy model, as implemented in the \Software{ViennaRNA}
package.  In the \Def{base pair model}, the total energy of a sequence $S$
for an RNA structure $R$ is given as sum of base pair energies, where
we consider six types of base pairs distinguishing by the bases A-U,
C-G or G-U (symmetrically) and between stacked and non-stacked base
pairs; here, we consider base pairs $(i,j)\in R$ \emph{stacked} iff
$(i+1,j-1)\in R$, otherwise \emph{non-stacked}. In the \Def{stacking model},
our features are defined by the stacks, i.e. pairs $(i,j)$ and
$(i+1,j-1)$ which both occur in $R$; we distinguish 18 types based on
$S_i,S_j,S_{i+1},S_{j-1}$ (i.e. all combinations that allow canonical
base pairs; the configurations $S_i,S_j,S_{i+1},S_{j-1}$ and
$S_{i+1},S_{j-1},S_j,S_i$ are symmetric).

>>>>

%\textcolor{red}{\textbf{SW: Simpler alternative to
%    Sec.~\ref{sec:first-improvedComplexity}. Note also that the use of
%    variables in Sec.~\ref{sec:first-improvedComplexity} conflicts
%    with the (recent) specialization of constraint networks to RNA design
%    networks.}}

@@

The correctness of the modified algorithm follows from the same
induction argument, where the message computation over one node
evaluates messages from its children only at cc-valid partial
sequences. The result of this computation is a message, which corresponds 
to the partition function, restricted to cc-valid partial sequences. 
Since invalid sequences have infinite energy, they do not contribute to 
the partition function, and the partition function restricted to cc-valid sequences coincides 
with the initial one.

This restriction drastically improves the time
complexity. Indeed, for any given node $v$, the original algorithm sends message for 
 $\chi(v):=\substitute{\val}{\val'}$ such that
$\val\in\partseqs(\separator{u}{v})$ and
$\val'\in\partseqs(\difference{u}{v})$, while the modified algorithm
only considers cc-valid assignments for $\chi(v)$. Remark that, in any connected component $cc$, 
assigning some nucleotide to a position reduces cc-valid assignments to (at most) 
two alternatives for each of the $|cc|-1$ remaining positions. It follows that, for a node $v$ featuring positions from $\#cc(v)$ distinct connected components $\{cc_1,cc_2\ldots\}$ in the base pair dependency graph, the number of cc-valid assignments to positions in $\chi(v)$ is exactly
$$4^{\#cc(v)}\prod_{i=1}^{\#cc(v)} 2^{|cc_i|-1} = 2^{\#cc(v)}\, 2^{|\chi(v)|} \in \Theta(2^{\#cc}\, 2^{w+1}),$$
for a single node, where $\#cc$ is the total number of connected components in the base pair dependency graph, and $w$ is the tree-width of the tree decomposition $T$. Since the number of nodes in $T$ is in $\Theta(n)$, and the number of atomic energy contributions associated with $k$ structures is in $\mathcal{O}(k\,n)$, then the overall complexity grows like $\Theta(n\, k\, 2^{\#cc}\, 2^{w+1})$.

%For each connected component having at least a position $i$ in $\chi(v)$, 
%the algorithm considers assigning any possible nucleotide $x\in \Sigma$ to $i$ and restricts
% the assignment of  for 
%any remaining position $j$ of the same connected component, it restricts the domain of $j$ to base-pairs that are consistent with the parity . 
%
%
%Let $c(v)$ be the number of connected components involving positions
%in $\chi(v)$; then, the first stage enumerates $2^{c(v)}$
%combinations. For each such combination, the second stage enumerates
%only $2^{|\chi(v)|}$ partial sequences, since there remain only two
%possible nucleotide choices for each position. All message values for
%node $v$ are thus computed in time
%$\Theta(2^{c(v)+|\chi(v)|})$. Consequently, the total time complexity
%of the modified algorithm is $$O(nk2^{\hat w}),$$ where $n$ is the RNA
%length, $k$ is the number of structures, and
%$\hat w := \max_{v\in T} c(v)+|\chi(v)|$.
%%
%From this directly follows the slightly looser complexity
%bound $$O(nk2^{w+1}2^{c})$$ in terms of treewidth $w$ and the maximum
%number $c$ of connected components represented in a node of the tree
%decomposition.  Notably, $\hat w$, as well as $w+1+c$, can not exceed
%$2(w+1)$, since $c\leq w+1$. Consequently the modified algorithm does
%never increase the original complexity.

Finally, we remark that even stronger time savings could be possible
in practice, since cc-valid partial sequences can still violate
complementarity constraints, e.g. by assigning C and A to positions in
different sets of a partition, thus satisfying the bi-partition
constraints, where the positions base pair directly, rendering the
partial sequence invalid. Moreover, applications of the sampling
framework, can introduce additional constraints that further reduce
the number of valid partial subsequences. However, exploiting all such
constraints, in a complete and general way, would likely cause significant
implementation overhead, while not significantly improving
the asymptotic complexity.

@@

We trained parameters for two RNA energy models to approximate the
Turner energy model, as implemented in the \Software{ViennaRNA}
package.  In the \Def{base pair model}, the total energy of a sequence $S$
for an RNA structure $R$ is given as sum of base pair energies, where
we consider six types of base pairs distinguishing by the bases $\Ab$-$\Ub$,
$\Cb$-$\Gb$ or $\Gb$-$\Ub$ (symmetrically) and between stacked and non-stacked base
pairs; here, we consider base pairs $(i,j)\in R$ \Def{stacked} iff
$(i+1,j-1)\in R$, otherwise \Def{non-stacked}. In the \Def{stacking model},
our features are defined by the stacks, i.e. pairs $(i,j)$ and
$(i+1,j-1)$ which both occur in $R$; we distinguish 18 types based on
$S_i,S_j,S_{i+1},S_{j-1}$ (i.e. all combinations that allow canonical
base pairs; the configurations $S_i,S_j,S_{i+1},S_{j-1}$ and
$S_{i+1},S_{j-1},S_j,S_i$ are symmetric).
